Combining neural networks and geostatistics for landslide hazard assessment of San Salvador Metropolitan Area, El Salvador \newline \newline Combinando redes neuronales y geoestad\'{i}stica para evaluaci\'{o}n de deslizamientos de tierra del \'{a}rea Metropolitana de San Salvador, El Salvador

### 2015-06-06 paper.tex

\section{Study area}
\label{sec:studyarea}
The Metropolitan Area of San Salvador (MASS), is formed by 14 municipalities among them there is San Salvador, the capital of the country. It has 587 km2 and according to the most recent census (\cite{minecon}) it has 1,566,629 inhabitans with a density of population of 2669 hab/km2.  According to geography MASS is extended over a flat erosional surface (650-760 m a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and with Mariona Hills (refer to Fig. \ref{fig:mass01}). Human settlement is not only restricted to the plain, but spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consists of intercaled primary and reworked deposits from Late Tertiary to Holocene age (\cite{schmidt1975}). The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in majority of MASS with an average thickness of 4 m (\cite{schmidt1975}). These are deposits derived mainly from Ilopango caldera eruptions where they present more than 50 m of thickness (\cite{schmidt1975}). Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking (\cite{schmidt1975}). Ch\'{a}vez et al. (\cite{chavez2014a}) identifies high and moderate erosion hazard in majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) (\cite{hernan2004}) present very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons), which results in high instability susceptibility (\cite{chavez2014b} and \cite{rolo2004}). Several incised rivers and ravines cross MASS area, due the large erosion rates of the area, natural slopes of these watercurses are often close to vertical and can reach highs of more than 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez (\cite{bommer}) identify in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hern\'{a}ndez (\cite{hernan2004}) carry out a detailed description of this process in TBJ, where these instabilities are very common. 
In study area fault trends are characterized by steep dip angles (65\textdegree - 90\textdegree). The main fault system is characterized by an east-west trend; this trend is responsible of the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting (\cite{schmidt1975}). However, all the mentioned fault families appear active. Thus the several fault systems have developed at different times, but been repeatedly activated (\cite{rymer1987}).   Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence (\cite{schmidt1975}).
Last eruption of San Salvador volcano was 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according to Sofield (\cite{sofield2004}). In the steep slopes  San Salvador volcano, especially of Picacho, the highest peak of the volcano, located in the eastern side present a high lahar hazard that threats MASS (\cite{major2004}). On the other hand one of the most destructive eruption in the world history was 430 A.D. TBJ eruption produced by the Ilopango Caldera. According to Dull (\cite{dull2004}), that eruption produced total devastation in whole MASS area.

\section{Landslide inventory and data sources of input variables}
\label{sec:landsinvet}
The Landslide inventory was developed by MARN and it was conducted in two stages. The first was carried out, based on fieldwork after the 2001 eartquakes in El Salvador. The another stage was completed with a visual inventory of places where there were landslides which were visible in a image of satellite image from IRS-C sensor (Indian) in the same year.

@@

Once that the landslide inventory was finished, a percentage of 0.5\% of the total points georeferenced shows landslide ocurrence. In addition to the landslide information the following data sources of input variables was provided by MARN:

@@

\item {\bf Slope:} Slope gradient is an important component and a
preparatory cause of landsliding (\cite{garcia2008}). It was calculated from digital elevation model (DEM) provided by MARN with a resolution of 10 mts.

\item {\bf Geology:} Description of the geology of the area from the map German Geological Mission (scale 1:100,000) (\cite{weber}).

\item {\bf Rainfall intensity:} From the precipitation database compiled by the
SNET's weather stations for the period 1970–2001, maximum precipitations were obtained, and then was applied interpolation kriging to them.

@@

\item {\bf Road distance:} Distance in kilometers to the nearest road. Road map was obtained from digitalized national cartography from Centro Nacional de Registros de la Rep\'{u}blica de El Salvador (CNR).

@@

\section{Artificial Neural Network model for discrete choice}
The logistic regression is a special case of neural network whose output variable is
discrete, logistic regression represents a neural network with 
a neuron in the hidden layer. The following adaptation of a multilayeredfeedforward artificial neural network known as  
MLP (Multilayer Perceptron) may be used for modeling binary classification model, where $x's$ are the observed values in the input variables, $w's$ and $\lambda's$ are the parameters of the model, $ p_{i} $ is the predicting probability for a network with $ k^{*} $ input characteristics and $ j^{*} $ neurons:

\begin{equation}
n_{j,i} = w_{j,0} + \sum_{k=1}^{k^{*}} w_{j,k}x_{k,i}
\end{equation}

\begin{equation}
N_{j,i} = \frac{1}{1+\exp^{-n_{j,i}}}
\end{equation}

\begin{equation}
p_{i} = \sum_{j=1}^{j^{*}} \lambda_{j} N_{j,i}
\end{equation}

\begin{equation}
\sum_{j=1}^{j^{*}} \lambda_{j} = 1 , \lambda_{j} \ge 0
\end{equation}

In the context of the research problem $p_{i} $ and the number $ k^{*} $ represent 
probability of landslide ocurrence and the number of input variables or causes associated with landslide ocurrence respectively.

@@

\begin{equation}
x*_{i} = \frac{x_{i} - \mu }{\sigma}
\label{eq:scaling}
\end{equation}

In the equation (\ref{eq:scaling}) $x*_{i}$ is the new variable transformed with $ \mu $ and $\sigma$ are, respectively the mean and the standard deviation for the ith input variable.

@@

Once a good estimate was obtained, this was occupied as the initial values for the conjugate gradient method implemented in the optim function of the R statistical software, to obtain a better estimation of the parameters of the model.

\section{Spatial Prediction}
In standard statistical problems, correlation can be estimated from a scatterplot, when several data pairs ${x, y}$ are available. The spatial correlation between two observations of a variable $z(s)$ at locations $s_{1}$ and $s_{2}$ cannot be estimated, as only a single pair is available. To estimate spatial correlation
from observational data, we therefore need to make stationarity assumptions
before we can make any progress. One commonly used form of stationarity
is intrinsic stationarity, which assumes that the process that generated the
samples is a random function $Z(s)$ composed of a mean and residual:

@@

It is a weighted average of the sample values, and $ \sum_{i=1}^{n} w_{i} = 1 $ to ensure unbiasedness. The $w_{i}$'s are the weights that will be estimated.

Kriging minimizes the mean squared error of prediction

@@

In order to make spatial predictions using ordinary kriging, a R script was developed which can be accessed in the following web address:

@@

In order to achieve the above steps a R script was developed which can be downloaded in the following address:

@@

>>>>

\section{Study area}
\label{sec:studyarea}
MASS, is formed by 14 municipalities among them it is located San Salvador, the capital of the country. It has 587 $km^{2}$ and according to the most recent census (\cite{minecon}) it has 1,566,629 inhabitans with a density of population of 2669 hab/km2.  According to geography MASS is extended over a flat erosional surface (650-760 m.a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and with Mariona Hills (refer to Fig. \ref{fig:mass01}). Human settlement is not only restricted to the plain, but spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consists of intercaled primary and reworked deposits from Late Tertiary to Holocene age (\cite{schmidt1975}). The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in majority of MASS with an average thickness of 4 m (\cite{schmidt1975}). These are deposits derived mainly from Ilopango caldera eruptions where they present more than 50 m of thickness (\cite{schmidt1975}). Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking (\cite{schmidt1975}). Ch\'{a}vez et al. (\cite{chavez2014a}) identifies high and moderate erosion hazard in majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) (\cite{hernan2004}) present very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons), which results in high instability susceptibility (\cite{chavez2014b} and \cite{rolo2004}). Several incised rivers and ravines cross MASS area, due the large erosion rates of the area, natural slopes of these watercourses are often close to vertical and can reach highs of more than 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez (\cite{bommer}) identified in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hern\'{a}ndez (\cite{hernan2004}) carried out a detailed description of this process in TBJ, where these instabilities are very common. 
In study area fault trends are characterized by steep dip angles (65\textdegree - 90\textdegree). The main fault system is characterized by an east-west trend; this trend is responsible of the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting (\cite{schmidt1975}). However, all the mentioned fault families appear active. Thus the several fault systems have developed at different times, but been repeatedly activated (\cite{rymer1987}).   Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence (\cite{schmidt1975}).
Last eruption of San Salvador volcano was 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according to Sofield (\cite{sofield2004}). In the steep slopes  San Salvador volcano, especially of Picacho, the highest peak of the volcano, located in the eastern side present a high lahar hazard that threats MASS (\cite{major2004}). On the other hand one of the most destructive eruptions in the world history was 430 A.D. TBJ eruption produced by the Ilopango Caldera. According to Dull (\cite{dull2004}), that eruption produced total devastation in whole MASS area.

\section{Landslide inventory and data sources of input variables}
\label{sec:landsinvet}
The Landslide inventory was developed by MARN and it was conducted in two stages. The first was carried out, based on fieldwork after the 2001 earthquakes in El Salvador. The second stage consisted of a visual inventory of places where there were landslides identified through satellite image from IRS-C sensor (Indian) in the same year.

@@

Once the landslide inventory was finished, a percentage of 0.5\% of the total points georeferenced showed landslide occurrence. In addition to the landslide information the following data sources of input variables was provided by MARN:

@@

\item {\bf Slope:} Slope gradient is an important component and a
preparatory cause of landsliding (\cite{garcia2008}). It was calculated from digital elevation model (DEM) provided by MARN with a resolution of 10 mt.

\item {\bf Geology:} Description of the geology of the area according digitalized geological map of  El Salvador (scale 1:100,000) (\cite{weber}).

\item {\bf Rainfall intensity:} From the precipitation database compiled by the
MARN's weather stations for the period 1970–2001, maximum precipitations were obtained, 
and then, a contour map was generated through kriging interpolation method.

@@

\item {\bf Road distance:} Distance in kilometers to the nearest road. Road map was obtained from digitalized national cartography created by Centro Nacional de Registros de la Rep\'{u}blica de El Salvador (CNR).

@@

\section{Artificial Neural Network model for discrete choice}
Logistic regression represents a neural network with 
a neuron in the hidden layer and whose output is dichotomous. The following adapted form of the Multilayer Perceptron (MLP) may be used for modeling binary classification problems. In the equation (\ref{eq:ann1}) $x_{k,i}$ are the observed values in the ith input variables belonging to the kth case. The expressions $w_{j,k}$ and $\lambda_{j}$ showed in the equations (\ref{eq:ann1}) and (\ref{eq:ann3}) are the parameters of the model.

\begin{equation}
n_{j,i} = w_{j,0} + \sum_{k=1}^{k^{*}} w_{j,k}x_{k,i}
\label{eq:ann1}
\end{equation}

\begin{equation}
N_{j,i} = \frac{1}{1+\exp^{-n_{j,i}}}
\label{eq:ann2}
\end{equation}

\begin{equation}
p_{i} = \sum_{j=1}^{j^{*}} \lambda_{j} N_{j,i}
\label{eq:ann3}
\end{equation}

\begin{equation}
\sum_{j=1}^{j^{*}} \lambda_{j} = 1 , \lambda_{j} \ge 0
\label{eq:ann4}
\end{equation}

In the equation (\ref{eq:ann3}) $p_{i}$ is the the predicting probability for a network with $ k^{*} $ input characteristics and $ j^{*} $ neurons. In the context of the present research, $p_{i} $ represents the probability of landslide ocurrence.

@@

\begin{equation}
z = \frac{x - \bar{x} }{\sigma_{x}}
\label{eq:scaling}
\end{equation}

In the equation (\ref{eq:scaling}) $z$ is the new variable transformed, $ \bar{x} $ and $\sigma_{x}$ are, respectively the mean and the standard deviation of the input variable to transform.

@@

Once a good estimate was obtained, this was occupied as the initial values for the conjugate gradient method implemented in the optim routine of the R statistical software, to obtain a better estimation of the parameters of the model.

\section{Spatial Prediction}

\subsection*{Estimating spatial correlation}

In standard statistical problems, correlation can be estimated from a scatterplot, when several data pairs ${x, y}$ are available. The spatial correlation between two observations of a variable $z(s)$ at locations $s_{1}$ and $s_{2}$ cannot be estimated, as only a single pair is available. To estimate spatial correlation
from observational data, it is necessary to make stationarity assumptions
before we can make any progress. One commonly used form of stationarity
is intrinsic stationarity, which assumes that the process that generated the
samples is a random function $Z(s)$ composed of a mean and residual \cite{bivand2008applied}:

@@

It is a weighted average of the sample values, and $ \sum_{i=1}^{n} w_{i} = 1 $ to ensure unbiasedness. The $w_{i}$'s are the weights that will be estimated. Kriging minimizes the mean squared error of prediction \cite{nicolas2015}.

@@

%In order to make spatial predictions using ordinary kriging, a R script was developed which can be accessed in the following web address: 
%
%\url{https://goo.gl/3N8n99}

@@

In order to achieve the above steps a R script was developed. This script can be downloaded in:

@@

\bibitem{bivand2008applied}
Bivand, Roger S and Pebesma, Edzer J and G{\'o}mez-Rubio, Virgilio and Pebesma, Edzer Jan
\newblock Applied spatial data analysis with R
\newblock \emph{volume 747248717}
\penalty0 Springer, 2014.

\bibitem{nicolas2015}
Nicolas Christou.
\newblock Ordinary kriging in terms of the covariance function
\newblock \emph{University of California, Los Angeles}
\penalty0 [cited 2015 June 04]. Available from http://goo.gl/0J1RkO
